---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# threshold_selection_paper

<!-- badges: start -->
<!-- badges: end -->

# Threshold selection for univariate extremes

R code to estimate a constant threshold, the excesses of which can be closely modelled by a Generalised Pareto distribution (GPD). 


## Motivation 

This method assumes data is independent and identically distributed. The function `thresh_qq_metric` selects a threshold from the set of proposed thresholds which allows for excesses to be modelled closely by a GPD while also accounting for parameter estimation uncertainty. 

When selecting a threshold for a particular dataset, there is a bias-variance trade-off which needs to be addressed. Too low a threshold is likely to violate the asymptotic basis of the GPD which would lead to bias in the fit of the model while choosing too high a threshold would generate very few exceedances with which the model can be estimated, leading to high parameter uncertainty. Thus, we want to choose as low a threshold as possible subject to the GPD providing a reasonable fit to the data. Using simulated data, our method has been shown to directly tackle this trade-off.


## Method Details

The threshold selection method modifies the approach developed in [Varty et al. (2021)](https://arxiv.org/abs/2102.00884), for use in a wider variety of extreme value modelling problems. Namely, the following adjustments have been made: 

	- observations are independent and identically distributed with continuous values,
	- a constant, rather than a variable, threshold is to be selected. 

The method has been shown to perform favourably in this context compared to competing methods. Further extensions are underway so that the method may be applied to more complicated settings, such as incorporating covariate dependence into threshold and/or GPD parameters.

### Mathematical Details

A threshold is selected from a user-specified set of candidate thresholds by using a quantile-based assessment to quantify the goodness-of-fit. Specifically, this is done by measuring the expected deviation from the line of equality on a QQ-plot for each candidate threshold.  

Let $y$ be the sample of excesses of some proposed threshold choice and $y^{(i)}$ be the $i^{\text{th}}$ bootstrapped sample of $y$. Denote by $Q_{(i)}(p) : [0,1] \rightarrow \mathbb{R}^+$ the sample quantile function of $y^{(i)}$ for $i = 1,\dots, k$ while $M_{(i)}(p) : [0,1] \rightarrow \mathbb{R}^+$ represents the model/theoretical quantile function based on the parameters fitted to $y^{(i)}$.

The metric, $d_{(i)}$, measures the disparity between the $i^{\text{th}}$ (of $k$) bootstrapped sample of threshold excesses and the relevant quantiles of the fitted GPD model. The discrepancy between the observed data and fitted model is evaluated at $m$ equally-spaced evaluation probabilities, $\{p_j = j / (m+1): j = 1,\dots,m\}$. This results in the following form for the threshold selection metric:

$$d_{(i)} = \frac{1}{m} \sum_{j=1}^{m} |M_{(i)}(p_j) - Q_{(i)}(p_j)|,$$

where $M_{(i)}(p) = \frac{\hat{\sigma}_i}{\hat{\xi}_i}\left[(1-p)^{-\hat{\xi}_i}-1\right]$ with $(\hat{\sigma}_i, \hat{\xi}_i)$ being the MLEs for the scale and shape parameters of the GPD fitted to the $i^{\text{th}}$ bootstrapped excesses $y^{(i)}$. 



The expected distance metric $d$, used to select the optimal threshold is then defined as: 

$$d = \frac{1}{k} \sum_{i = 1}^{k} d_{(i)}.$$

## Example Usage 

```{r, echo=TRUE}
source("helper_functions.R")
source("thresh_qq_metric.R")
set.seed(12345)
data_test1 <- rgpd(1000, shape = 0.1, scale=0.5, mu=1)
thresholds1 <- seq(0.5, 2.5, by=0.1)
example1 <- thresh_qq_metric(data_test1, thresh = thresholds1)
example1
plot(thresholds1, example1$dists, xlab="Threshold", ylab="Metric value")

set.seed(11111)
test2 <- rgpd(10000, shape = 0.1, scale=0.5)
u <- 1
cens_thr<-u*rbeta(length(test2),1,0.5)
keep <- test2>cens_thr
data_test2 <- test2[keep]
thresholds2 <- quantile(data_test2,seq(0, 0.95, by=0.05))
example2 <- thresh_qq_metric(data_test2,thresh = thresholds2)
example2
```
